---
params:
  set_title: "new title"
  set_author: "new author"
  showCode: FALSE
  ifPlatePlot: TRUE
title: "`r params$set_title`"
author: "`r params$set_author`"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---

# Dataset summary

```{r, echo = FALSE}
library(knitr)
library(kableExtra)
library(gridExtra)
library(DrugScreenExplorer)
library(tidyverse)
#read in data, will not show.
screenData <- readRDS("screenData.rds")
```

## Basic information

### Screen format
```{r screen format, echo=params$showCode, warning=FALSE, message=FALSE}
screenInfo <- tibble(info = c("Plate Number", "Plate format"), 
                    value = c(length(unique(screenData$fileName)), length(unique(screenData$wellID))))

if ("batch" %in% names(screenData)) screenInfo <- rbind(screenInfo, c("Batch number",length(unique(screenData$batch))))
if ("name" %in% names(screenData)) screenInfo <- rbind(screenInfo, c("Number of drugs/probs",length(unique(screenData$name))))
if ("wellType" %in% names(screenData)) screenInfo <- rbind(screenInfo, c("Well types",paste0(unique(screenData$wellType), collapse = ",")))

kable(screenInfo, align = "c", col.names = NULL, format = "html") %>% kable_styling()
```

### Plate layout

```{r prepare layout, echo=params$showCode, warning=FALSE, message=FALSE}
if (! "plateID" %in% colnames(screenData)) screenData$plateID <- 1
plotTab <- filter(screenData, !duplicated(paste0(wellID,"_",plateID))) %>% 
  mutate(fileName = sprintf("layout %s",plateID))
plotList <- platePlot(plotTab, plotType = "layout")
nLayout <- length(unique(plotTab$plateID))
```


```{r plot layout, echo=params$showCode, fig.width=10, fig.height=((nLayout+1) %/% 2)*3}
grid.arrange(grobs = plotList, ncol=2)
```

## Raw signal distribution on each plate
```{r plot raw distribution, echo=params$showCode, warning=FALSE, fig.height=8, fig.width=8}
#check how many plates are there in the screen
nPlate <- length(unique(screenData))
g <- plotRawCount(screenData, ifLog10=TRUE)

if (nPlate <= 20) {
  plot(g)
} else {
  gTop20 <- plotRawCount(screenData, topN = 20, ifLog10 = TRUE)
  plot(gTop20)
  
  #create a pdf file for all plates
  pdf("plots/rawCountDistribution.pdf", width = 8, height = nPlate*0.7)
  plot(g)
  dev.off()
}
```
`r if(nPlate > 20) {"Only the first 20 plates are shown in html report. The plot for all plates can be found [here](./plots/rawCountDistribution.pdf)"}`


`r if(params$ifPlatePlot) {"## Plate plot"}`

```{r prepare plate plot, echo= params$showCode & params$ifPlatePlot, eval = params$ifPlatePlot}
if (!"normVal" %in% colnames(screenData)) {
  print("Normalized viablity not found. Using z-score")
  g <- platePlot(screenData, plotType = "zscore")
} else {
  print("Normalized viablity found. Using viability values")
  g <- platePlot(screenData, plotType = "viability")
}

nPlate <- length(unique(screenData$fileName))
```

```{r plot plate plots, echo= params$showCode & params$ifPlatePlot, eval = params$ifPlatePlot, fig.width=16, fig.height=((nPlate+1) %/% 2)*3}
if (nPlate <= 10) {
  grid.arrange(grobs = g, ncol = 2)
} else {
  grid.arrange(grobs = g[1:10], ncol = 2)
  makepdf(g, "plots/platePlots.pdf", 2, 3, 12, 12)
}
```
`r if(nPlate > 10 & params$ifPlatePlot) {"Only the first 10 plates are shown in html report. The plot for all plates can be found [here](./plots/platePlots.pdf)"}`


# Quality assessment

## Variance vs rank of mean

```{r, echo=params$showCode, fig.width=8, fig.height=8}
perPlateVar <- mutate(screenData, log10Val = log10(value)) %>% group_by(fileName) %>%
  summarise(meanRaw = mean(value), sdRaw = sd(value),
            meanLog = mean(log10Val), sdLog = sd(log10Val))

customTheme = theme_bw() + theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p1 <- arrange(perPlateVar, meanRaw) %>% mutate(rank = seq(nrow(perPlateVar))) %>%
  ggplot(aes(x=rank, y = sdRaw)) + geom_point(color = "royalblue") + xlab("rank (mean)") + ylab("standard deviation") +
  ggtitle("Per-plate") + customTheme
p2 <- arrange(perPlateVar, meanLog) %>% mutate(rank = seq(nrow(perPlateVar))) %>%
  ggplot(aes(x=rank, y = sdLog)) + geom_point(color = "royalblue") + xlab("rank (mean)") + ylab("standard deviation") +
  ggtitle("Per-plate (log10 scale)") + customTheme

perWellVar <- mutate(screenData, log10Val = log10(value)) %>% group_by(wellID) %>%
  summarise(meanRaw = mean(value), sdRaw = sd(value),
            meanLog = mean(log10Val), sdLog = sd(log10Val))

p3 <- arrange(perWellVar, meanRaw) %>% mutate(rank = seq(nrow(perWellVar))) %>%
  ggplot(aes(x=rank, y = sdRaw)) + geom_point(color = "firebrick") + xlab("rank (mean)") + ylab("standard deviation") +
  ggtitle("Per-well") + customTheme
p4 <- arrange(perWellVar, meanLog) %>% mutate(rank = seq(nrow(perWellVar))) %>%
  ggplot(aes(x=rank, y = sdLog)) + geom_point(color = "firebrick") + xlab("rank (mean)") + ylab("standard deviation") +
  ggtitle("Per-well (log10 scale)") + customTheme

grid.arrange(p1,p2,p3,p4, ncol = 2)
```

## Edge effect

### Plot intensity of edge effect
```{r prepare edge effect table, echo=params$showCode}
nLayer <- 3
edgePlot <- estimateEdgeEffect(screenData, nLayer = nLayer, onlyNeg = TRUE) %>%
  select(value, fileName, paste0("edgeRatio",seq(nLayer))) %>%
  filter(!duplicated(fileName))

rankByMean <- group_by(screenData, fileName) %>% summarise(rawMean = mean(value)) %>%
  arrange(rawMean)

mapCol <- structure(seq(nLayer), names = paste0("edgeRatio", seq(nLayer)))
plotTab <- arrange(edgePlot, value) %>% mutate(fileName = factor(fileName, levels = rev(rankByMean$fileName))) %>%
  select(fileName, names(mapCol)) %>% gather(key = "layer", value = "edgeRatio", -fileName) %>% 
  mutate(layer = factor(mapCol[layer])) %>% group_by(fileName, layer) %>% 
  summarise(meanFac = mean(edgeRatio), sdFac = sd(edgeRatio)) 


```

```{r plot edge intensity, echo = params$showCode}
#create plot
p <- ggplot(plotTab, aes(x=fileName, y = meanFac, fill = layer)) +
 geom_bar(position = position_dodge(), stat = "identity") + coord_flip() + theme_bw() +
 xlab("file names (ranked by mean of raw counts)") + ylab("edge effect ratio") + scale_fill_brewer(palette = 1)

allNames <- unique(plotTab$fileName)
if (length(allNames) > 10) {
  plotTab_top20 <- filter(plotTab, fileName %in% allNames[1:10])
  p_top20 <- ggplot(plotTab_top20, aes(x=fileName, y = meanFac, fill = layer)) + 
    geom_bar(position = position_dodge(), stat = "identity") + coord_flip() + theme_bw() + 
    xlab("file names (ranked by mean of raw counts)") + ylab("edge effect ratio") + scale_fill_brewer(palette = 1)
  plot(p_top20)
  pdf("plots/edgeEffectIntensity.pdf", width = 10, height = length(allNames)*0.6)
  plot(p)
  dev.off()
} else {
  plot(p)
}
```
`r if(length(allNames) > 10) {"Only the first 20 plates are shown in html report. The plot for all plates can be found [here](./plots/edgeEffectIntensity.pdf)."}`


### PCA plot
```{r PCA, echo = params$showCode, fig.height=10, fig.width=10}
#Wether this data has been normlized by netagive controls?
if (! "normVal" %in% colnames(screenData)) {
  #no normalization 
  print("No viability data found, using per-plate z-score")
  calcZ <- function(x) mutate(x, normVal = (value-mean(value))/sd(value))
  matPlate <- group_by(screenData, fileName) %>% do(calcZ(.))
} else matPlate <- screenData

#prepare matrix for PCA
matPlate <- select(matPlate, wellID, fileName, value) %>% spread(key = wellID, value = value) %>% 
  data.frame()
fileNameList <- matPlate$fileName
matPlate$fileName <- NULL
matPlate <- log2(as.matrix(matPlate))

#perform PCA
pcRes <- prcomp(matPlate, center= TRUE, scale. = FALSE)
varExp <- pcRes$sdev^2
varExp <- varExp/sum(varExp)
plotTab <- tibble(PC1 = pcRes$x[,1], PC2= pcRes$x[,2],
                  fileName = fileNameList)

#add batch information if present
if ("batch" %in% colnames(screenData)) {
  plotTab <- mutate(plotTab, batch = factor(screenData[match(fileNameList, screenData$fileName),]$batch))
  p <- ggplot(plotTab, aes(x=PC1, y = PC2, col = batch, label = fileName)) + geom_point()
} else {
  p <- ggplot(plotTab, aes(x=PC1, y = PC2, label = fileName)) + geom_point(col = "royalblue")
}
p <- p + theme_bw() + xlab(sprintf("PC1 (%1.2f%%)", varExp[1]*100)) + ylab(sprintf("PC2 (%1.2f%%)", varExp[2]*100)) +
  ggtitle("Principal component analysis on plates") +  geom_text(size=3, nudge_y = -0.1) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
plot(p)
```

### MDS
```{r MDS, echo = params$showCode, fig.height=10, fig.width=10}
mdsRes <- cmdscale(dist(matPlate))
plotTab <- tibble(X = mdsRes[,1], Y= mdsRes[,2],
                  fileName = fileNameList)
#add batch information if present
if ("batch" %in% colnames(screenData)) {
  plotTab <- mutate(plotTab, batch = factor(screenData[match(fileNameList, screenData$fileName),]$batch))
  p <- ggplot(plotTab, aes(x=X, y = Y, col = batch, label = fileName)) + geom_point()
} else {
  p <- ggplot(plotTab, aes(x=X, y = Y, label = fileName)) + geom_point(col = "royalblue")
}

p <- p + theme_bw() + ggtitle("Multidimensional Scaling (MDS) on plates") +  geom_text(size=3, nudge_y = -0.1) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
plot(p)
```


